Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
add ñ

---
documentclass: jss
author:
  - name: Hanne I. Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Munoz"
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands \AND
  - name: Thomas P. A. Debray
    affiliation: Julius Center for Health Sciences and Primary Care
    address: |
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands   
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Valentijn M. T. de Jong
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
    affiliation2: |
      | Data Analytics and Methods Task Force, 
      | European Medicines Agency, Amsterdam, The Netherlands

>>>>

---
documentclass: jss
author:
  - name: Hanne I. Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Muñoz"
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands \AND
  - name: Thomas P. A. Debray
    affiliation: Julius Center for Health Sciences and Primary Care
    address: |
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands   
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Valentijn M. T. de Jong
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
    affiliation2: |
      | Data Analytics and Methods Task Force, 
      | European Medicines Agency, Amsterdam, The Netherlands

### 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
add ñ

\author{
Hanne I. Oberman\\Utrecht University \And Johanna Munoz\\Julius Center
for Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
for Health Sciences and Primary Care \And Gerko Vink\\Utrecht University
\AND Valentijn M. T. de Jong\\Julius Center for Health Sciences and
Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne I. Oberman, Johanna Munoz, Thomas P. A. Debray, Gerko
Vink, Valentijn M. T. de Jong}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{Multilevel \pkg{mice}}

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     4.808     0.197    24.354     5.072     0.000     7.930     0.916 
sex             1.086     0.135     8.056     5.271     0.000     6.761     0.902

                            Estimate 
Intercept~~Intercept|school    0.324 
Residual~~Residual             1.130 
ICC|school                     0.226

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     4.808     0.197    24.354     5.072     0.000     7.930     0.916 
sex             1.086     0.135     8.056     5.271     0.000     6.761     0.902

                            Estimate 
Intercept~~Intercept|school    0.324 
Residual~~Residual             1.130 
ICC|school                     0.226

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- imp %>% 
+   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
R> tidy(pool(fit))
\end{CodeInput}
\begin{CodeOutput}
                     term   estimate   std.error  statistic      p.value
1             (Intercept) -2.3054756 0.340854445  -6.763813 1.425704e-11
2                    type -0.4200458 0.180837517  -2.322780 2.021049e-02
3                     age  0.0302518 0.001561625  19.372002 0.000000e+00
4 as.factor(motor_score)2 -0.6619736 0.069527977  -9.520967 0.000000e+00
5 as.factor(motor_score)3 -1.0520025 0.070748986 -14.869507 0.000000e+00
6 as.factor(motor_score)4 -1.5087772 0.072240455 -20.885489 0.000000e+00
7                   pupil  0.4731045 0.038187004  12.389150 0.000000e+00
8                      ct  0.4291135 0.032126389  13.357041 0.000000e+00
             b          df dfcom         fmi      lambda m         riv
1 7.937082e-04  9227.60708 11013 0.008412824 0.008197930 5 0.008265691
2 1.386594e-04 10229.67142 11013 0.005282534 0.005088076 5 0.005114097
3 1.138007e-08 10083.78177 11013 0.005796971 0.005599802 5 0.005631337
4 1.427540e-04  2450.43742 11013 0.036222753 0.035436457 5 0.036738334
5 1.147119e-04  3540.28145 11013 0.028049975 0.027501049 5 0.028278744
6 1.377409e-04  2902.11217 11013 0.032339200 0.031672563 5 0.032708526
7 3.160434e-04    58.71181 11013 0.284053948 0.260073900 5 0.351486317
8 1.985834e-04    74.37520 11013 0.250767587 0.230887509 5 0.300199921
          ubar
1 1.152293e-01
2 3.253582e-02
3 2.425016e-06
4 4.662835e-03
5 4.867765e-03
6 5.053394e-03
7 1.078995e-03
8 7.938047e-04
\end{CodeOutput}
\begin{CodeInput}
R> as.mitml.result(fit)
\end{CodeInput}
\begin{CodeOutput}
[[1]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10494.825 10560.593 -5238.412 10476.825     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2917  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.33693                 -0.43143                  0.03013  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.67764                 -1.05120                 -1.49989  
                  pupil                       ct  
                0.49336                  0.44123  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings

[[2]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10543.607 10609.376 -5262.803 10525.607     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2937  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.28159                 -0.42088                  0.03026  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65610                 -1.04928                 -1.49432  
                  pupil                       ct  
                0.48055                  0.41014  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[3]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10543.170 10608.939 -5262.585 10525.170     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2901  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.29948                 -0.40181                  0.03038  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65959                 -1.06391                 -1.52075  
                  pupil                       ct  
                0.45837                  0.42186  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings

[[4]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10527.222 10592.991 -5254.611 10509.222     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2962  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.27660                 -0.42912                  0.03016  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.66963                 -1.05952                 -1.51971  
                  pupil                       ct  
                0.45085                  0.44439  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[5]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10519.178 10584.947 -5250.589 10501.178     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2897  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.33278                 -0.41699                  0.03032  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.64691                 -1.03611                 -1.50921  
                  pupil                       ct  
                0.48240                  0.42794  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

>>>>

\author{
Hanne I. Oberman\\Utrecht University \And Johanna Muñoz\\Julius Center
for Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
for Health Sciences and Primary Care \And Gerko Vink\\Utrecht University
\AND Valentijn M. T. de Jong\\Julius Center for Health Sciences and
Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne I. Oberman, Johanna Muñoz, Thomas P. A. Debray, Gerko
Vink, Valentijn M. T. de Jong}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{Multilevel \pkg{mice}}

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     4.915     0.217    22.642     4.926     0.000     9.110     0.926 
sex             0.975     0.283     3.444     4.250     0.024    32.504     0.978

                            Estimate 
Intercept~~Intercept|school    0.351 
Residual~~Residual             1.153 
ICC|school                     0.233

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     5.011     0.410    12.222     4.226     0.000    35.955     0.980 
sex             0.928     0.381     2.434     4.168     0.069    48.221     0.985

                            Estimate 
Intercept~~Intercept|school    0.313 
Residual~~Residual             1.428 
ICC|school                     0.188

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- imp %>% 
+   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
R> tidy(pool(fit))
\end{CodeInput}
\begin{CodeOutput}
                     term    estimate   std.error  statistic      p.value
1             (Intercept) -2.35203726 0.340181747  -6.914061 4.994037e-12
2                    type -0.41265892 0.180274846  -2.289054 2.209524e-02
3                     age  0.03049023 0.001570162  19.418521 1.238416e-81
4 as.factor(motor_score)2 -0.66764920 0.068737865  -9.712975 3.480413e-22
5 as.factor(motor_score)3 -1.05520001 0.070218940 -15.027285 2.540225e-50
6 as.factor(motor_score)4 -1.51238349 0.072304262 -20.916934 1.850073e-90
7                   pupil  0.48421447 0.038982800  12.421234 6.772069e-17
8                      ct  0.43474621 0.029968474  14.506785 2.342253e-36
             b          df dfcom         fmi      lambda m         riv
1 5.281599e-04 10119.70041 11013 0.005673265 0.005476771 5 0.005506932
2 4.881699e-05 10893.89378 11013 0.001985736 0.001802528 5 0.001805783
3 3.335358e-08  6320.89582 11013 0.016545467 0.016234340 5 0.016502243
4 4.188703e-05  8327.24771 11013 0.010875748 0.010638213 5 0.010752602
5 5.135721e-05  7632.20045 11013 0.012757637 0.012498967 5 0.012657168
6 1.403058e-04  2831.75462 11013 0.032888241 0.032205434 5 0.033277139
7 3.571497e-04    49.97295 11013 0.309130930 0.282023647 5 0.392803531
8 8.586645e-05   294.69765 11013 0.120677044 0.114729598 5 0.129598366
          ubar
1 1.150898e-01
2 3.244044e-02
3 2.425385e-06
4 4.674630e-03
5 4.869071e-03
6 5.059539e-03
7 1.091079e-03
8 7.950697e-04
\end{CodeOutput}
\begin{CodeInput}
R> as.mitml.result(fit)
\end{CodeInput}
\begin{CodeOutput}
[[1]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10495.423 10561.192 -5238.712 10477.423     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2843  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.37195                 -0.41014                  0.03052  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65802                 -1.04611                 -1.51245  
                  pupil                       ct  
                0.50405                  0.42496  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[2]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
     AIC      BIC   logLik deviance df.resid 
10500.88 10566.65 -5241.44 10482.88    11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2917  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.37718                 -0.41511                  0.03067  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.66935                 -1.05211                 -1.49429  
                  pupil                       ct  
                0.49013                  0.43835  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[3]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10505.026 10570.795 -5243.513 10487.026     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2908  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.32339                 -0.42359                  0.03023  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.67142                 -1.05776                 -1.51038  
                  pupil                       ct  
                0.49756                  0.42474  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[4]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10519.511 10585.280 -5250.755 10501.511     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2961  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.33581                 -0.40871                  0.03039  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.66477                 -1.05451                 -1.51858  
                  pupil                       ct  
                0.45928                  0.44419  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings

[[5]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10522.038 10587.807 -5252.019 10504.038     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2955  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.35187                 -0.40574                  0.03064  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.67468                 -1.06551                 -1.52622  
                  pupil                       ct  
                0.47006                  0.44148  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
